Effectiveness of COVID-19 Booster doses in England: a sequential trials study in OpenSAFELY

### 2022-02-14 manuscript/draft-manuscript.Rmd

hazards_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "effects.csv")) %>%
  mutate(
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
    hr_display = paste0(number_format(accuracy=0.01)(hr), " (", number_format(accuracy=0.01)(hr.ll), "-", number_format(accuracy=0.01)(hr.ul), ")")
  )

@@

A total of `r filter(matchsummary, treatment=="pfizer") %>% pull(n_treated))` BNT162b2 booster doses were administered between `r format(as.Date(study_dates$studystart_date), "%d %B %Y")` and `r format(as.Date(study_dates$studyend_date), "%d %B %Y")` in adults registered at a TPP practice, with `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_eligible))` (`r label_percent(0.1)(filter(matchsummary, treatment=="pfizer") %>% pull(prop_eligible_treated))`) eligible for inclusion into the treatment arm (supplementary Tables S1 and S2). Of these, `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` (`r label_percent(0.1)(filter(matchsummary, treatment=="pfizer") %>% pull(prop_matched_eligible))`) were successfully matched with unboosted controls (Figure 1).

<!-- For mRNA-1273, this was `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="moderna") %>% pull(n))` (`r label_percent(0.1)(filter(matchsummary, treatment=="moderna") %>% pull(prop_recruited))`) and `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="moderna") %>% pull(fup_years))` person-years of follow-up (Table 2).  -->

@@

xmin <- min(matchcoverage$vax3_date )
xmax <- max(matchcoverage$vax3_date )+1
matchcoverage  %>%
  mutate(
    # vax12_type = fct_case_when(
    #   vax12_type == "pfizer-pfizer" ~ "BNT162b2",
    #   vax12_type == "az-az" ~ "ChAdOx1-S",
    #   vax12_type == "moderna-moderna" ~ "mRNA-1273",
    #   TRUE ~ NA_character_
    # ),
    course = fct_case_when(
      vax12_type=="pfizer-pfizer" & treatment=="pfizer" ~ "2x BNT162b2  / 1x BNT162b2",
      vax12_type=="az-az" & treatment=="pfizer" ~ "2 x ChAdOx1-S / 1 x BNT162b2",
      vax12_type=="pfizer-pfizer" & treatment=="moderna" ~ "2x BNT162b2 / 1x mRNA-1273",
      vax12_type=="az-az" & treatment=="moderna" ~ "2x ChAdOx1-S / 1x mRNA-1273",
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot() +
  geom_col(
    aes(
      x=vax3_date+0.5,
      y=cumuln,
      group=status,
      fill=status,
      colour=NULL
    ),
    position=position_stack(),
    alpha=0.8,
    width=1
  )+
  geom_rect(xmin=xmin, xmax= xmax+1, ymin=0, ymax=6, fill="grey", colour="transparent")+
  facet_wrap(vars(course))+
  scale_x_date(
    breaks = unique(lubridate::ceiling_date(matchcoverage$vax3_date, "1 month")),
    limits = c(lubridate::floor_date((xmin), "1 month"), NA),
    labels = scales::label_date("%d/%m"),
    expand = expansion(add=1),
  )+
  scale_y_continuous(
    labels = scales::label_number(accuracy = 1, big.mark=","),
    expand = expansion(c(0, NA))
  )+
  scale_fill_brewer(type="qual", palette="Set2")+
  scale_colour_brewer(type="qual", palette="Set2")+
  labs(
    x="Date",
    y="Cumulative booster vaccines",
    colour=NULL,
    fill=NULL,
    alpha=NULL
  ) +
  theme_minimal()+
  theme(
    axis.line.x.bottom = element_line(),
    axis.text.x.top=element_text(hjust=0),
    strip.text.y.right = element_text(angle = 0),
    axis.ticks.x=element_line(),
    legend.position = "bottom"
  )+
  NULL
cat("  \n\n")
```

@@

```{r table1, echo=FALSE, warning=FALSE, message=FALSE}
table1_body <- read_csv(fs::path(output_dir_os, "matchtable1.csv"))
table1_by <- read_csv(fs::path(output_dir_os, "matchtable1by.csv"))

@@

```{r table2, echo=FALSE, warning=FALSE, message=FALSE}
incidences %>%
  filter(
    fup_period == "All",
    outcome %in% c("postest", "covidadmitted", "coviddeath")
  ) %>%
  left_join(
    matchsummary %>% select(treatment, eligible_1=total_treated, recruit_prop_1 = prop_recruited),
    by=c("treatment")
  ) %>%
  select(
    treatment_descr, 
    outcome_descr,
    eligible_1,
    n_1,
    recruit_prop_1,
    q_1,
    rate_1,
    q_0, 
    rate_0,
    rr, 
    rrCI
  ) %>%
  gt(
    groupname_col = c("treatment_descr"),
  ) %>%
  cols_label(
    #treatment_descr = "Treatment",
    outcome_descr = "Outcome",

@@

<!-- For mRNA-1273, the corresponding estimated hazard ratios were `r filter(hazards, outcome=="postest", treatment=="moderna", model==3, term_left==28) %>% pull(hr_display)`, for COVID-19 A&E attendances was `r filter(hazards, outcome=="covidattendance", treatment=="moderna", model==3, term_left==28) %>% pull(hr_display)`, for COVID-19 hospital admission was `r filter(hazards, outcome=="covidadmitted", treatment=="moderna", term_left==28) %>% pull(hr_display)`, for COVID-19 critical care admission was `r filter(hazards, outcome=="covidcc", treatment=="moderna", model==3, term_left==28) %>% pull(hr_display)`, and for COVID-19 death was `r filter(hazards, outcome=="coviddeath", treatment=="moderna", model==3, term_left==28) %>% pull(hr_display)`.  -->

For the same period, the estimated effectiveness for BNT162b2 and ChAdOx1-S primary courses respectively were  `r filter(hazards_vax12_type, outcome=="postest", treatment=="pfizer", vax12_type=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="postest", treatment=="pfizer", vax12_type=="az-az", model==3, term_left==28) %>% pull(hr_display)` for a positive test, `r filter(hazards_vax12_type, outcome=="covidadmitted", treatment=="pfizer", vax12_type=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="covidadmitted", treatment=="pfizer", vax12_type=="az-az", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 hospital admission, and `r filter(hazards_vax12_type, outcome=="coviddeath", treatment=="pfizer", vax12_type=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="coviddeath", treatment=="pfizer", vax12_type=="az-az", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 death (Figure 2).

<!--  `r filter(hazards_vax12_type, outcome=="covidemergency", treatment=="pfizer", vax12_type=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="emergency", treatment=="pfizer", vax12_type=="az-az", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 A&E attendances, -->
<!-- `r filter(hazards_vax12_type, outcome=="covidcc", treatment=="pfizer", vax12_type=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="covidcc", treatment=="pfizer", vax12_type=="az-az", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 critical care admission, -->

@@

  plot_hazards <-
    hazards3_vax12_type %>%
    filter(outcome %in% c("postest", "covidadmitted", "coviddeath")) %>%
    ggplot()+
    geom_point(aes(y=hr, x=term_midpoint, colour=model_descr), position = position_dodge(width = 1.8))+
    geom_linerange(aes(ymin=hr.ll, ymax=hr.ul, x=term_midpoint, colour=model_descr), position = position_dodge(width = 1.8))+
    geom_hline(aes(yintercept=1), colour='grey')+
    facet_grid(rows=vars(outcome_descr), cols=vars(subgroup_descr), switch="y")+
    scale_y_log10(
      breaks=y_breaks,
      limits = c(0.01, 1.5),
      oob = scales::oob_keep,
      sec.axis = sec_axis(
        ~(1-.),
        name="Effectiveness",
        breaks = 1-(y_breaks),
        labels = function(x){formatpercent100(x, 1)}
      )
    )+
    scale_x_continuous(breaks=postbaselinecuts, limits=c(min(postbaselinecuts), max(postbaselinecuts)+1), expand = c(0, 0))+
    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=1))+
    labs(
      y="Hazard ratio",
      x="Days since booster dose",
      colour=NULL
     ) +
    theme_bw()+
    theme(
      panel.border = element_blank(),
      axis.line.y = element_line(colour = "black"),
      axis.line.y.right = element_blank(),

@@

This observational study of `r label_number(1, big.mark=",")(flowchart[[nrow(flowchart), "n"]]*2)` fully vaccinated people living in England estimated high protective effects of booster vaccination against all studied outcomes. severe COVID-19 disease and death. Protection appeared to be highest in those receiving a X-X-X course, though . There were apparent waning effects after X weeks which is unlikely to explained by increased transmissibility of the Omicron variant given the low prevalence of Omicron at the end of the follow-up window.

### Strengths and weaknesses

We used routinely-collected health records with comprehensive coverage of primary care, hospital admissions, COVID-19 testing, COVID-19 vaccination, and death registrations to study fully vaccinated English residents. ...

The methodological approach used is robust but inefficient; many boosted individuals are not retained after matching, and those who are retained are censored when their matched control is boosted. Millions of eligible participants may be required to obtain precise estimates, particularly when event rates are low, and so extending the approach to smaller subgroups may not be feasible. FOr instance, due to small numbers, we did not study booster effectiveness in those who had received the mRNA-1273 vaccine as their primary course or those who had received a heterologous primary course.

@@

@@

@@

[pfizer phase 3 booster trial, in pfizer first course]
[ cov-boost trial]

>>>>

hazards_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "effects.csv")) %>%
  mutate(
    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level_descr = fct_inorder(subgroup_level_descr),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
    hr_display = paste0(number_format(accuracy=0.01)(hr), " (", number_format(accuracy=0.01)(hr.ll), "-", number_format(accuracy=0.01)(hr.ul), ")")
  )

@@

A total of `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_treated))` BNT162b2 booster doses were administered between `r format(as.Date(study_dates$studystart_date), "%d %B %Y")` and `r format(as.Date(study_dates$studyend_date), "%d %B %Y")` in adults registered at a TPP practice, with `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_eligible))` (`r label_percent(0.1)(filter(matchsummary, treatment=="pfizer") %>% pull(prop_eligible_treated))`) eligible for inclusion into the treatment arm. Of these, `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` (`r label_percent(0.1)(filter(matchsummary, treatment=="pfizer") %>% pull(prop_matched_eligible))`) were successfully matched with unboosted controls (Figure 1).

<!-- For mRNA-1273, this was `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="moderna") %>% pull(n))` (`r label_percent(0.1)(filter(matchsummary, treatment=="moderna") %>% pull(prop_matched_eligible))`) and `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="moderna") %>% pull(fup_years))` person-years of follow-up (Table 2).  -->

@@

xmin <- min(matchcoverage$vax3_date )
xmax <- max(matchcoverage$vax3_date )+1
matchcoverage  %>%
  filter(treatment=="pfizer") %>%
  mutate(
    # vax12_type = fct_case_when(
    #   vax12_type == "pfizer-pfizer" ~ "BNT162b2",
    #   vax12_type == "az-az" ~ "ChAdOx1-S",
    #   vax12_type == "moderna-moderna" ~ "mRNA-1273",
    #   TRUE ~ NA_character_
    # ),
    course = fct_case_when(
      vax12_type=="pfizer-pfizer" & treatment=="pfizer" ~ "2x BNT162b2  / 1x BNT162b2",
      vax12_type=="az-az" & treatment=="pfizer" ~ "2 x ChAdOx1-S / 1 x BNT162b2",
      vax12_type=="pfizer-pfizer" & treatment=="moderna" ~ "2x BNT162b2 / 1x mRNA-1273",
      vax12_type=="az-az" & treatment=="moderna" ~ "2x ChAdOx1-S / 1x mRNA-1273",
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot() +
  geom_col(
    aes(
      x=vax3_date+0.5,
      y=cumuln,
      group=fct_rev(status),
      fill=fct_rev(status),
      colour=NULL
    ),
    position=position_stack(),
    alpha=0.8,
    width=1
  )+
  geom_rect(xmin=xmin, xmax= xmax+1, ymin=0, ymax=6, fill="grey", colour="transparent")+
  facet_wrap(vars(course))+
  scale_x_date(
    breaks = unique(lubridate::ceiling_date(matchcoverage$vax3_date, "1 month")),
    limits = c(lubridate::floor_date((xmin), "1 month"), NA),
    labels = scales::label_date("%d/%m"),
    expand = expansion(add=1),
  )+
  scale_y_continuous(
    labels = scales::label_number(accuracy = 1, big.mark=","),
    expand = expansion(c(0, NA))
  )+
  scale_fill_brewer(type="qual", palette="Set2", direction = -1)+
  scale_colour_brewer(type="qual", palette="Set2", direction = -1)+
  labs(
    x="Date",
    y="Cumulative booster vaccines",
    colour=NULL,
    fill=NULL,
    alpha=NULL
  ) +
  theme_minimal()+
  theme(
    axis.line.x.bottom = element_line(),
    axis.text.x.top=element_text(hjust=0),
    strip.text.y.right = element_text(angle = 0),
    axis.ticks.x=element_line(),
    legend.position = "bottom"
  )+
  NULL
cat("  \n\n")
```

@@

```{r table1, echo=FALSE, warning=FALSE, message=FALSE}
table1_body <- read_csv(fs::path(output_dir_os, "matchtable1.csv")) %>% filter(treatment=="pfizer")
table1_by <- read_csv(fs::path(output_dir_os, "matchtable1by.csv")) %>% filter(treatment=="pfizer")

@@

```{r table2, echo=FALSE, warning=FALSE, message=FALSE}
incidences %>%
  filter(
    fup_period == "All",
    outcome %in% c("postest", "covidadmitted", "coviddeath")
  ) %>%
  left_join(
    matchsummary %>% select(treatment, eligible_1=n_eligible, recruit_prop_1 = prop_matched_eligible),
    by=c("treatment")
  ) %>%
  select(
    treatment_descr, 
    outcome_descr,
    eligible_1,
    n_1,
    recruit_prop_1,
    q_1,
    rate_1,
    q_0, 
    rate_0,
    rr, 
    rrCI
  ) %>%
  gt(
    groupname_col = c("treatment_descr"),
  ) %>%
  cols_label(
    #treatment_descr = "Treatment",
    outcome_descr = "Outcome",

@@

<!-- For mRNA-1273, the corresponding estimated hazard ratios were `r filter(hazards_none, outcome=="postest", treatment=="moderna", model==3, term_left==28) %>% pull(hr_display)`, for COVID-19 A&E attendances was `r filter(hazards_none, outcome=="covidattendance", treatment=="moderna", model==3, term_left==28) %>% pull(hr_display)`, for COVID-19 hospital admission was `r filter(hazards_none, outcome=="covidadmitted", treatment=="moderna", term_left==28) %>% pull(hr_display)`, for COVID-19 critical care admission was `r filter(hazards_none, outcome=="covidcc", treatment=="moderna", model==3, term_left==28) %>% pull(hr_display)`, and for COVID-19 death was `r filter(hazards_none, outcome=="coviddeath", treatment=="moderna", model==3, term_left==28) %>% pull(hr_display)`.  -->

For the same period, the estimated effectiveness for BNT162b2 and ChAdOx1-S primary courses respectively were `r filter(hazards_vax12_type, outcome=="postest", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="postest", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==28) %>% pull(hr_display)` for a positive test, `r filter(hazards_vax12_type, outcome=="covidadmitted", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="covidadmitted", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 hospital admission, and `r filter(hazards_vax12_type, outcome=="coviddeath", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="coviddeath", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 death (Figure 2).

@@

  plot_hazards <-
    hazards3_vax12_type %>%
    filter(outcome %in% c("postest", "covidadmitted", "coviddeath")) %>%
    ggplot()+
    geom_point(aes(y=hr, x=term_midpoint, colour=model_descr), position = position_dodge(width = 1.8))+
    geom_linerange(aes(ymin=hr.ll, ymax=hr.ul, x=term_midpoint, colour=model_descr), position = position_dodge(width = 1.8))+
    geom_hline(aes(yintercept=1), colour='grey')+
    facet_grid(rows=vars(outcome_descr), cols=vars(subgroup_level_descr), switch="y")+
    scale_y_log10(
      breaks=y_breaks,
      limits = c(0.01, 1.5),
      oob = scales::oob_keep,
      sec.axis = sec_axis(
        ~(1-.),
        name="Effectiveness",
        breaks = 1-(y_breaks),
        labels = function(x){formatpercent100(x, 1)}
      )
    )+
    scale_x_continuous(breaks=postbaselinecuts, limits=c(min(postbaselinecuts), max(postbaselinecuts)+1), expand = c(0, 0))+
    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=1))+
    labs(
      y="Hazard ratio",
      x="Days since booster dose",
      colour=NULL
     ) +
    theme_bw()+
    theme(
      panel.border = element_blank(),
      axis.line.y = element_line(colour = "black"),
      axis.line.y.right = element_blank(),

@@

This observational study of fully vaccinated adults in England estimated high protection of BNT162b2 booster vaccination against all studied outcomes. 
Protection appeared higher in those receiving a X-X primary course.
There were apparent waning effects after X weeks which is unlikely to explained by increased transmissibility of the Omicron variant given Omicron was not yet dominant in the UK by the end of the follow-up window (ref).

### Limitations

We used routinely-collected health records with comprehensive coverage of primary care, hospital admissions, COVID-19 testing and vaccination, and death registrations to study over X million fully vaccinated English residents ...

@@

Positive SARS-CoV-2 testing data likely underestimates the true incidence of infection. The mass availability of lateral flow tests in the UK during the study period, whose results are not routinely recorded in COVID-19 surveillance data, means many infections, including symptomatic infections, may be undocumented, despite encouragement to seek a confirmatory PCR test. Potential differences in testing behaviour between vaccinated and unvaccinated people also contributes to the unreliability of testing data as a means to assess effectiveness.

@@

The methodological approach used is robust but inefficient; many boosted individuals are not retained after matching, and those who are retained are censored when their matched control is boosted. Millions of eligible participants may be required to obtain precise estimates, particularly when event rates are low, and so extending the approach to smaller subgroups may not be feasible. For instance, due to small numbers, we did not study booster effectiveness in those who had received the mRNA-1273 vaccine as their primary course or those who had received a heterologous primary course, despite many thousands of such recipients in our records.

Finally, mRNA-1273 boosting was rolled out later than for BNT162b2, and so we could not evaluate effectiveness of this vaccine due to insufficient follow-up. A analysis of mRNA-1273 boosting and comparative effectiveness between mRNA-1273 and BNT162b2 is planned.

@@

[pfizer phase 3 booster trial, in pfizer first course], [ cov-boost trial]

A phase III trial assessing BNT162b2 booster efficacy in two-dose BNT162b2 recipients found.. (ref)
